Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-28 numerical.tex
urgent revision

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$, except with the possible emergence
of modes associated with boundary conditions. Thus, tracking the
fundamental mode is a more preferable assessment of stability as $\beta$
is increased from zero, rather than searching for the most unstable
mode. This is because the latter may correspond to modes that depend on boundary
conditions, e.g. surface modes, whose growth rate varies with
$\zmax$.

@@

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
We now test the critical thermal relaxation timescale given by
Eq. \ref{iso_vsi_cond}. We begin from eigenvalues for the fundamental
mode with $\beta=10^{-3}$ and slowly increase the thermal
relaxation timescale up to $\beta=10^3$. Fig. \ref{bcrit_compare1}
show the obtained growth rates, together with the critical  
thermal relaxation timescale $\beta_\mathrm{crit}=0.125$ for our
fiducial disk parameters.

@@

 Fig. \ref{bcrit_compare}, together with Fig. \ref{bcrit_compare1} and 
 our finding that the fundamental mode becomes dominant as one
 increases $\beta$, and that it is not sensitive to boundary
 conditions, shows that $\beta_\mathrm{crit}$ as given  by
 Eq. \ref{iso_vsi_cond} provides a practical measure for the thermal
 relaxation timescale below which  VSI will operate.

@@

\subsection{Vertically non-isothermal disks} 
We briefly consider vertically non-isothermal disks with 
$\Gamma=1.4$ and $(p,q,\epsilon)=(0,-1,0.05)$ as simulated in
\cite{nelson13}. In this case we set $\zmax=0.99H_s$.

Fig. \ref{gcorr_compare_vnoniso} plots the fundamental VSI growth
rates as a function of $\beta$ for $\gamma\in[1.4,2.5]$. In agreement
with \citeauthor{nelson13}, in   
the neutrally-stratified case $\gamma=\Gamma$ the disk is unstable 
even for $\beta\gg 1$. This is due to the absence of a stabilizing
vertical entropy gradient ($N_z^2\equiv 0$). % (Note that in the adiabatic limit 
% this disk can be unstable according to the
% Solberg-Hoiland criterion, Eq. \ref{solberg2}.)

For $\gamma>\Gamma$, i.e. stably stratified disks,
Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
relaxation rapidly stabilizes the disk, similar to that observed for
nearly vertically isothermal disks (Fig. \ref{bcrit_compare1}). For
$\gamma=1.7,\,2.0,\,2.5$, growth rates reach zero at
$\beta\simeq0.17,\,0.083,\,0.045$, respectively. Interestingly, these
values are equal to $\epsilon|q|/(\gamma-\Gamma)$. This suggests that
the critical thermal relaxation timescale for vertically
non-isothermal disks can also be estimated by Eq. \ref{iso_vsi_cond}
but with $\gamma-1$ replaced by $\gamma-\Gamma$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/gcorr_compare_vnoniso2}
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation timescale $\beta$, in vertically
    non-isothermal disks with $\Gamma=1.4$ and
    $\gamma\in[1.4,2.5]$. The disk is neutrally
    stratified for $\gamma=1.4$ and stably stratified for
    $\gamma>1.4$. Other disk parameters are
    $(p,q,\epsilon)=(0,-1,0.05)$ and the perturbation wavenumber is
    $\khat=30$.   
    \label{gcorr_compare_vnoniso}}
\end{figure}

>>>>

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$, except with the possible emergence
of modes associated with boundary conditions. Thus, tracking the
fundamental mode is a more preferable assessment of stability as $\beta$
is increased from zero, rather than searching for the most unstable
mode. This is because the latter may correspond to modes that depend on boundary
conditions, e.g. surface modes whose growth rate varies with
$\zmax$ in a vertically isothermal disk.

@@

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
We now test the critical thermal relaxation timescale given by
Eq. \ref{iso_vsi_cond}. We iterate starting from eigenvalues for the fundamental
mode with $\beta=10^{-3}$ and slowly increase the thermal
relaxation timescale up to $\beta=10^3$. Fig. \ref{bcrit_compare1}
show the growth rates obtained this way, together with the critical  
thermal relaxation timescale $\beta_\mathrm{crit}=0.125$ for our
fiducial disk parameters.

@@

 Fig. \ref{bcrit_compare}, together with Fig. \ref{bcrit_compare1} and 
 our finding that the fundamental mode becomes dominant as one
 increases $\beta$, and that it is not sensitive to boundary
 conditions, shows that $\beta_\mathrm{crit}$ as given  by
 Eq. \ref{iso_vsi_cond} provides a practical measure for the thermal
 relaxation timescale below which  VSI will operate in a vertically
 isothermal disk.

@@

\subsection{Vertically non-isothermal disks}

% We briefly consider vertically non-isothermal disks with 
% $\Gamma=1.4$ and $(p,q,\epsilon)=(0,-1,0.05)$ as simulated in
% \cite{nelson13}. In this case we set $\zmax=0.99H_s$.

% Fig. \ref{gcorr_compare_vnoniso} plots the fundamental VSI growth
% rates as a function of $\beta$ for $\gamma\in[1.4,2.5]$. In agreement
% with \citeauthor{nelson13}, in   
% the neutrally-stratified case $\gamma=\Gamma$ the disk is unstable 
% even for $\beta\gg 1$. This is due to the absence of a stabilizing
% vertical entropy gradient ($N_z^2\equiv 0$). % (Note that in the adiabatic limit 
% % this disk can be unstable according to the
% % Solberg-Hoiland criterion, Eq. \ref{solberg2}.)

% For $\gamma>\Gamma$, i.e. stably stratified disks,
% Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
% relaxation rapidly stabilizes the disk, similar to that observed for
% nearly vertically isothermal disks (Fig. \ref{bcrit_compare1}). For
% $\gamma=1.7,\,2.0,\,2.5$, growth rates reach zero at
% $\beta\simeq0.17,\,0.083,\,0.045$, respectively. Interestingly, these
% values are equal to $\epsilon|q|/(\gamma-\Gamma)$. This suggests that
% the critical thermal relaxation timescale for vertically
% non-isothermal disks can also be estimated by Eq. \ref{iso_vsi_cond}
% but with $\gamma-1$ replaced by $\gamma-\Gamma$.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   0cm]{figures/gcorr_compare_vnoniso2}
%   \caption{Growth rate of the fundamental VSI mode as a function of
%     the thermal relaxation timescale $\beta$, in vertically
%     non-isothermal disks with $\Gamma=1.4$ and
%     $\gamma\in[1.4,2.5]$. The disk is neutrally
%     stratified for $\gamma=1.4$ and stably stratified for
%     $\gamma>1.4$. Other disk parameters are
%     $(p,q,\epsilon)=(0,-1,0.05)$ and the perturbation wavenumber is
%     $\khat=30$.   
%     \label{gcorr_compare_vnoniso}}
% \end{figure}
